---
title: "R Code"
author: "Grace, Irazu"
date: "2025-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sunscreen Toxicity Assay Data Analysis

-   by Mary Grace Catapang, Irazu Casado

```{r}
library(dplyr)
library(dunn.test)
library(ggplot2)
library(ecotox)
```

## 1. Data cleaning and preparation

```{r load data}
data <- read.csv("sunscreen_data3.csv")
data
```
```{r}
# Split by time and apply correction separately
data_24 <- subset(data, Time_hr == 24)
data_24 <- abbott_correction(data_24)

data_48 <- subset(data, Time_hr == 48)
data_48 <- abbott_correction(data_48)
```


## 2. Analyzing Control Data

No significant difference between the Control and DMSO groups: \* 24hr - all recorded 0 deaths for both Control and DMSO \* 48hr - nonparametric Wilcox test show non-significant results (p = 0.8137)

```{r}
control48data <- data %>% filter(ControlType == "Control") %>% filter(Time_hr == 48)
control48data
dmso48data <- data %>% filter(ControlType == "DMSO") %>% filter(Time_hr == 48)
dmso48data
```

```{r}
wilcox.test(control48data$MortalityRate, dmso48data$MortalityRate, alternative="two.sided") 
```

Because there is no significant difference, we can use both Control and DMSO as pooled control.

```{r}
# Create a pooled control variable
data$PooledControl <- ifelse(data$ControlType %in% c("Control", "DMSO"), "PooledControl", "Treatment")
data$PooledControl <- as.factor(data$PooledControl)
data
```



## 3. Calculations of NOEC and LOEC using Kruskal-Wallis + Dunn's test

No significant difference in the mortality rates of different concentrations

```{r}
# Check normality
shapiro.test(data$MortalityRate[data$PooledControl== "PooledControl"])
shapiro.test(data$MortalityRate[data$PooledControl== "Treatment"])
```

Not normally distributed.

```{r}
library(car)
leveneTest(MortalityRate ~ PooledControl, data = data)
```

p \< 0.05, variances are unequal

### NOEC / LOEC at 48hr

Kruskal Wallis test across concentrations (including pooled controls)

```{r}
kw48 <- kruskal.test(MortalityRate ~ Concentration_sunscreen, data = data_48)
kw48
```
```{r}
# Post-hoc test 
dunn48 <- dunn.test(data_48$MortalityRate, data_48$Concentration_sunscreen, method = "bh")
```

### NOEC / LOEC at 24hr 

Kruskal Wallis test across concentrations (including pooled controls)

```{r}

kw24 <- kruskal.test(MortalityRate ~ Concentration_sunscreen, data = data_24)
kw24
```

```{r}
# Post-hoc test 
dunn24 <- dunn.test(data_24$MortalityRate, as.factor(data_24$Concentration_sunscreen), method = "bh")
```



### COMPARATION 24H and 48H per group of concentraction:

```{r}

resultados_wilcox <- data %>%
  group_by(Concentration_sunscreen) %>%
  summarise(
    p_value = tryCatch(
      wilcox.test(
        MortalityRate[Time_hr == 24],
        MortalityRate[Time_hr == 48],
        paired = TRUE,
        exact = FALSE
      )$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  )

print(resultados_wilcox)


```

### Mortality across different concentrations / original data

```{r}
library(ggplot2)
library(dplyr)


summary_data <- data %>%
  group_by(Concentration_percent, Time_hr) %>%
  summarise(
    mean_mortality = mean(MortalityRate),
    se = sd(MortalityRate) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    ci95 = se * qt(0.975, df = n - 1),
    Time_hr = factor(Time_hr)
  )

ggplot(summary_data, aes(x = factor(Concentration_percent),
                         y = mean_mortality,
                         fill = Time_hr)) +
  geom_bar(stat = "identity", position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_mortality - ci95,
                    ymax = mean_mortality + ci95),
                width = 0.2,
                position = position_dodge(0.8)) +
  scale_fill_manual(
    values = c("24" = "#1f77b4", "48" = "#ff7f0e"),
    labels = c("24 h", "48 h"),
    name = "Exposure time"
  ) +
  labs(
    title = "Mean Mortality ± 95% CI by Concentration and Time",
    x = expression(bold("Concentration (%)")),
    y = expression(bold("Proportional Mortality (0–1)"))
  ) +
  theme_classic(base_family = "serif", base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(size = 14),
    legend.position = "top",
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 13),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid = element_blank()
  )

```

## 4. LC50 values

calcula intervalo confianza

```{r function}
get_LC_values <- function(df_subset) {
  # Remove zero or negative concentrations to avoid log10(0) or log10(negative)
  df_subset <- df_subset %>% filter(Concentration_sunscreen > 0)
  
  # Check for missing/infinite values
  df_subset <- df_subset %>% filter(is.finite(Concentration_sunscreen),
                                    is.finite(Dead),
                                    is.finite(NumberExposed))
  
  # Check that there’s variation in mortality
  if (length(unique(df_subset$Dead / df_subset$NumberExposed)) == 1) {
    stop("No variation in mortality - cannot fit GLM")
  }
  
  # Fit the probit GLM
  model <- glm(cbind(Dead, NumberExposed - Dead) ~ log10(Concentration_sunscreen),
                data = df_subset, family = binomial(link = "probit"))
  
  # Helper function to calculate LCx
  get_LC <- function(model, x) {
    z <- qnorm(x / 100)
    coef <- coef(model)
    logLC <- (z - coef[1]) / coef[2]
    10^logLC
  }
  
  # Calculate LC10 and LC50
  lc10 <- get_LC(model, 10)
  lc50 <- get_LC(model, 50)
  
  # Return the values and model
  return(list(LC10 = lc10, LC50 = lc50, model = model))
}

```

### LC50 at 24h


```{r}
get_LC_values(data_24)
```

Convert to mg/L, considering stock solution is 1g/mL

```{r}
# Example LC values (decimal)
LC_values <- c(0.002876023, 0.01877322)  # LC10 and LC50 in decimal

# Conversion to mg/L
mg_per_L <- LC_values * 10000

# Create a data frame with both decimal and mg/L values
df_LC <- data.frame(
  LC_decimal = LC_values,
  LC_mg_per_L = mg_per_L
)

# Print results
print(df_LC)
```

### LC50 at 48h

```{r}
get_LC_values(data_48)
```

```{r}
# Example LC values (decimal)
LC_values_48 <- c(0.001207872 , 0.008641191)  # LC10 and LC50 in decimal

# Conversion to mg/L
mg_per_L <- LC_values_48 * 10000

# Create a data frame with both decimal and mg/L values
df_LC <- data.frame(
  LC_decimal = LC_values_48,
  LC_mg_per_L = mg_per_L
)

# Print results
print(df_LC)
```

## 5. Probit dose-response curve

```{r}
if (!any(grepl("h", data$Time_hr))) {
  data$Time_hr <- factor(data$Time_hr,
                                    levels = c(24, 48),
                                    labels = c("24h", "48h"))
}
data <- data %>% filter(PooledControl=="Treatment")
data <- data %>% filter(Concentration_sunscreen> 0)
data$Alive <- data$NumberExposed - data$Dead
data$logConc <- log10(data$Concentration_percent)
data$MortalityRate <- data$Dead / data$NumberExposed

data_24h <- subset(data, Time_hr == "24h")
data_48h <- subset(data, Time_hr == "48h")

# Ajustar modelos Probit
model_probit_24h <- glm(cbind(Dead, Alive) ~ logConc, data = data_24h, family = binomial(link = "probit"))
model_probit_48h <- glm(cbind(Dead, Alive) ~ logConc, data = data_48h, family = binomial(link = "probit"))

# Generar secuencia de concentraciones logarítmicas para predicción
log_conc_seq <- seq(min(data$logConc), max(data$logConc), length.out = 100)

# Crear dataframes de predicción
pred_24h <- data.frame(logConc = log_conc_seq, Time_hr = "24h")
pred_24h$Mortality <- predict(model_probit_24h, newdata = pred_24h, type = "response")

pred_48h <- data.frame(logConc = log_conc_seq, Time_hr = "48h")
pred_48h$Mortality <- predict(model_probit_48h, newdata = pred_48h, type = "response")

# Unir predicciones
pred_all <- rbind(pred_24h, pred_48h)

# Crear el gráfico
ggplot() +
  geom_point(data = data,
             aes(x = logConc, y = MortalityRate, color = Time_hr),
             size = 3, alpha = 0.7) +
  geom_line(data = pred_all,
            aes(x = logConc, y = Mortality, color = Time_hr),
            linewidth = 1.2) +
  scale_color_manual(values = c("24h" = "#1f77b4", "48h" = "#ff7f0e"),
                     labels = c("24h", "48h")) +
  theme_minimal(base_size = 12) +
  labs(title = "Probit Dose-Response Curve",
       x = "Log10(Concentration Percent)",
       y = "Mortality proportion",
       color = "Exposition time") +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12))
```

```{r}
library(ggplot2)

ggplot() +
  geom_point(
    data = data,
    aes(x = logConc, y = MortalityRate, color = Time_hr, shape = Time_hr),
    size = 3, alpha = 0.8
  ) +
  geom_line(
    data = pred_all,
    aes(x = logConc, y = Mortality, color = Time_hr),
    linewidth = 1.3
  ) +
  scale_color_manual(
    values = c("24h" = "#1f77b4", "48h" = "#ff7f0e"),
    labels = c("24 h", "48 h")
  ) +
  scale_shape_manual(
    values = c(16, 17),
    labels = c("24 h", "48 h")
  ) +
  labs(
    title = "Probit Dose-Response Curve",
    x = expression(bold("log"[10]*"(Concentration %)")),
    y = expression(bold("Mortality proportion")),
    color = "Exposure time",
    shape = "Exposure time"
  ) +
  theme_classic(base_family = "sans", base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid = element_blank()
  )

```

